---
title: "Assignment 4"
author: "Sahil Ganatra"
date: "2025-03-23"
output: html_document
---

```{r}
if (!require('BiocManager', quietly = TRUE))
    install.packages('BiocManager')

BiocManager::install('GenomicFeatures')
BiocManager::install("TxDb.Hsapiens.UCSC.hg19.knownGene")
```

```{r}
library(tidyverse)                        # data manipulation and visualization
library(rtracklayer)                      # importing genomic files
library(GenomicRanges)                    # for handling genomic intervals
library(GenomicFeatures)                  # for working with gene models
library(TxDb.Hsapiens.UCSC.hg19.knownGene)    # human genome annotations
```

```{r}
# list of BigWig files from ATAC-seq experiment
# the experiment compares 2 clones (Clone3 and Clone16) w/ and w/o IAA treatment
# IAA treatment depletes CTCF protein through the auxin-inducible degron system
bw_files <- c("GSM6072254_JWH213-C16MinusHudep-1963139.free.bw", # Clone16 control
"GSM6072252_JWH211-C3MinusHudep-1963137.free.bw",       # Clone3 control
"GSM6072255_JWH214-C16PlusHudep-1963140.free.bw",       # Clone16 IAA treated
"GSM6072253_JWH212-C3PlusHudep-1963138.free.bw")        # Clone3 IAA treated

# Import each BigWig file into a GRanges object
gr_list <- lapply(bw_files, function(file) {
import(file, format = "BigWig")
})

# Name each element in the list for easy reference
names(gr_list) <- c("Clone16_Untreated", "Clone3_Untreated",
"Clone16_IAA", "Clone3_IAA")

# Display summary of the first dataset
summary(gr_list[[1]])

Clone16_Untreated <- length(gr_list[["Clone16_Untreated"]])
Clone16_IAA <- length(gr_list[["Clone16_IAA"]])

# print the counts
cat("Clone 16 Untreated has: ", Clone16_Untreated, "ranges\n")
cat("Clone 16 IAA has: ", Clone16_IAA, "ranges\n")

counts <- c(Clone16_Untreated, Clone16_IAA)
barplot(counts,
        main ='Number of Ranges in Clone16 Conditions',
        ylab = 'Number of Ranges',
        col = c('lightblue', 'lightgreen'))
```



Questions: 

1) What are the core components of a GRanges object? List them.

GRanges is a vector of genomic locations and associate annotations. Each element
in the vector consists of a sequence name, an interval, and a strand.

2) Compare the number of ranges between Clone16_Untreated and Clone16_IAA.

Clone16_Untreated has 1828285 ranges, and Clone16_IAA has 1549060 ranges. This
indicates that IAA treatment results in fewer regions of open chormatin being
detected.


```{r}
# Extracting Clone16 Untreated data for analysis
gr1 <- gr_list[["Clone16_Untreated"]]

# Demonstrating basic range operations
# 1. Subset ranges on specific chromosomes
chr1_ranges <- gr1[seqnames(gr1) == "chr1"]

# 2. Find ranges within a specific genomic window
region_of_interest <- GRanges("chr1", IRanges(1, 1000000)) # First 1Mb of chr1
window_peaks <- subsetByOverlaps(gr1, region_of_interest)

# 3. Calculate coverage
peak_coverage <- coverage(gr1)

# Calculate basic statistics
chr1_stats <- data.frame(
mean_score = mean(chr1_ranges$score),
median_score = median(chr1_ranges$score),
n_peaks = length(chr1_ranges)
)
```

Questions (un-comment and fill in the code):
```{r}
# 1) Find all ranges on chromosome 2:
chr2_ranges <- gr1[seqnames(gr1) == 'chr2']

# 2) Calculate the median score:
median_score <- median(chr2_ranges$score)
```

Part 3: Promoter Analysis
```{r}
# Load gene model database for human genome (hg19)
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene

# Extract all gene coordinates
genes <- genes(txdb)
```
```{r}
# Define promoter regions (±2kb around transcription start sites)
# The promoters() function automatically handles strand information
promoter_regions <- promoters(genes, upstream = 2000, downstream = 2000)
seqlevelsStyle(promoter_regions) <- seqlevelsStyle(gr_list[[1]])
promoter_regions
```
```{r}
# Demonstrate range arithmetic
# reduce() merges overlapping ranges
promoter_reduced <- reduce(promoter_regions)

# disjoin() creates non-overlapping ranges
promoter_disjoin <- disjoin(promoter_regions)

# Find closest genes to peaks
nearest_genes <- nearest(gr_list[[1]], promoter_regions)
```

Questions:

1) Why do we use promoters() instead of just adding/subtracting from gene start
   positions?

The promoters() function automatically accounts for the strand information
of each gene. Simply adding or subtracting a fixed number from the gene start
positions would ignore that genes on the negative strand have their 
transcription start sites oriented in the opposite direction.

2) What does the seqlevelsStyle() function do and why is it necessary?

The seqlevelsStyle() function allows us to get/set the naming convention for the
sequence levels (chromosome names) in the GRanges objects. This is necessary 
because different genomic resources may use different naming conventions.

Part 4: Gene Annotation & Overlap Analysis
```{r}
# Load human gene annotation database
library(org.Hs.eg.db)
```
```{r}
# Extract Entrez gene IDs and map to gene symbols
entrez_ids <- as.character(promoter_regions$gene_id)
gene_symbols <- mapIds(org.Hs.eg.db,
                       keys = entrez_ids,
                       column = 'SYMBOL',       # get gene symbols
                       keytype = 'ENTREZID',    # input ID type
                       multiVals = 'first')     # take first mapping if multiple exist
```
```{r}
# Add gene symbols to promoter regions
promoter_regions$gene_symbol <- gene_symbols

# Find overlaps between ATAC-seq peaks and promoter regions
overlap_results <- lapply(gr_list, function(gr) {
  overlaps <- findOverlaps(gr, promoter_regions)
  gr[queryHits(overlaps)]
})

# Calculate overlap statistics
overlap_stats <- lapply(overlap_results, function(gr) {
  data.frame(
    n_peaks = length(gr),
    mean_score = mean(gr$score),
    median_score = median(gr$score)
  )
})

overlap_stats
```

Questions:

1) What does findOverlaps() return?

The findOverlaps() function compares 2 GRanges objects and returns indices of 
intervals that overlap between them. It's used to determine which peaks from 
gr1 fall within defined genomic regions.

2) Uncomment and fill in the blanks in the code below to find only overlaps on
   chromosome 1:
```{r}
chr1_overlaps <- findOverlaps(
  gr_list[[1]][seqnames(gr_list[[1]]) == 'chr1'],
  gr_list[[2]][seqnames(gr_list[[2]]) == 'chr1']
)
```

Part 5: Differential Accessibility Analysis
```{r}
library(pheatmap)
library(ggplot2)

# Function to analyze differential accessibility
analyze_diff_accessibility <- function(gr_list, promoter_regions, top_n = 20) {
  # Option 1: Filter out promoter regions with missing gene symbols
  promoter_regions <- promoter_regions[!is.na(promoter_regions$gene_symbol)]
  
  # Create accessibility matrix
  promoter_matrix <- matrix(0,
                            nrow = length(promoter_regions),
                            ncol = length(gr_list))
  
  # Add names to matrix (make gene symbols unique and convert to character)
  rownames(promoter_matrix) <- make.unique(as.character(promoter_regions$gene_symbol))
  colnames(promoter_matrix) <- names(gr_list)
  
  # Fill in matrix with accessibility scores
  for (i in seq_along(gr_list)) {
    overlaps <- findOverlaps(gr_list[[i]], promoter_regions)
    if (length(overlaps) > 0) {
      # Sum scores for each promoter
      scores <- split(gr_list[[i]]$score[queryHits(overlaps)],
                      subjectHits(overlaps))
      scores_sum <- sapply(scores, sum)
      promoter_matrix[as.numeric(names(scores_sum)), i] <- scores_sum
    }
  }
  
  # Remove rows with all zeros or NA values
  valid_rows <- rowSums(promoter_matrix) > 0 &
    !apply(promoter_matrix, 1, function(x) any(is.na(x)))
  promoter_matrix <- promoter_matrix[valid_rows, ]
  
  # Calculate fold changes
  fold_changes <- data.frame(
    gene_symbol = rownames(promoter_matrix),
    Clone16_FC = log2((promoter_matrix[, 'Clone16_IAA'] + 1) /
                      (promoter_matrix[, 'Clone16_Untreated'] + 1)),
    Clone3_FC = log2((promoter_matrix[, 'Clone3_IAA'] + 1) / 
                      (promoter_matrix[, 'Clone3_Untreated'] + 1))
  )
  
  # Calculate average absolute fold change
  fold_changes$avg_abs_FC <- (abs(fold_changes$Clone16_FC) +
                              abs(fold_changes$Clone3_FC)) / 2
  
  # Select top differential promoters
  top_genes <- fold_changes[order(fold_changes$Clone16_FC, decreasing = TRUE), 
                            ][1:min(top_n, nrow(fold_changes)), ]
  
  # Create heatmap using as.character() to ensure correct subsetting
  top_matrix <- promoter_matrix[as.character(top_genes$gene_symbol), ]
  pheatmap(top_matrix,
           main = 'Top Differentially Accessible Promoters',
           color = colorRampPalette(c('white', 'navy', 'red'))(100),
           cluster_rows = TRUE,
           cluster_cols = TRUE,
           fontsize_row = 9,
           fontsize_col = 10)
  
  # Create scatter plot
  scatter <- ggplot(fold_changes, aes(x = Clone3_FC, y = Clone16_FC)) + 
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'darkgray') +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'darkgray') + 
  geom_text(data = top_genes, aes(label = gene_symbol),
            size = 3, hjust = -0.2, vjust = -0.2) +
  labs(title = 'Comparison of IAA vs. Untreated Accessibility Changes',
       x = 'Log2 Fold Change in Clone 3',
       y = 'Log2 Fold Change in Clone 16') +
  theme_minimal()

  print(scatter)

  return(list(fold_changes = fold_changes, top_genes = top_genes))
}

# Run differential accessibility analysis
results <- analyze_diff_accessibility(gr_list, promoter_regions)
results$top_genes %>% head()
```


Questions - Looking at the heatmap and scatter plot:

1) What patterns do you observe in the differential accessibility between 
   IAA and untreated conditions?

The heatmap shows that many genes in both Clone3 and Clone16 become more 
accessible when treated with IAA, but this is more prevalent with Clone16. The 
scatter plot shows that most promoters cluster around (0, 0), indicating no 
large difference bwtween IAA and untreated in either clone. However, there are 
some promoters far from the origin in the upper-right quadrant, which is 
consistent with the heatmap showig strong increases in accessibility for certain
promoters upon IAA treatment.

2) Why might we use both a heatmap and a scatter plot to visualize this data?

While the scatterplot makes it easier to assess overall trends and to identify 
outliers, the heatmap allows us to see clusters or patterns that might group 
promoters with similar behavior. Together, these two visualizations offer both a
broad overview and a more focused comparison of differential changes between 
conditions.

Part 6: Trying It Yourself
```{r}
ctcf_promoter <- promoter_regions[!is.na(promoter_regions$gene_symbol) 
                                  & promoter_regions$gene_symbol == 'CTCF']
ctcf_window <- resize(ctcf_promoter, width(ctcf_promoter) + 10000, fix='center')
nearby_peaks <- gr1[queryHits(findOverlaps(gr1, ctcf_window))]

# count the number of peaks within the extended CTCF promoter windows
num_ctcf_peaks <- length(nearby_peaks)
print(num_ctcf_peaks)
```

1) How many peaks fall within 5kb of CTCF promoters?

Running the code above produced 244 peaks that fall within 5kb of CTCF 
promoters.
